---
title: "Notes for Intro To R Carpentry"
author: "Aleksandra Wilczynska"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE , 
                      message = FALSE
                      )
```

# Before we Start
## Why R?
<div class="alert alert-info">
<strong>Go to slides</strong> 
</div>

## Why Rstudio
<div class="alert alert-info">
<strong>Go to slides</strong> 
</div>

### Tour around R Studio
<div class="alert alert-info">
<strong>Go to slides</strong> 
</div>

**Show:**

1. Panes 

2. Working directory and project **create a project**: `data-carpentry`

3. Create directories 
  - `data`
  - `data_output`
  - `documents`
  - `fig_output`
  - `scripts`

4. Explain difference between the console and the script (chat vs email)

5. Create a script: `script.R`  - save in the `scripts` folder 

6. Explain shortcuts: 
    - `CTRL + ENTER` or `Cmd + Return` 
    
7. Installing packages 
    - Check if packages are installed ( interface )
    - Install packages (interface + console)

```{r eval = F}
install.packages('tidyverse')
```

8. Downloading files
```{r eval = F}
download.file("https://ndownloader.figshare.com/files/11492171",
              "data/SAFI_clean.csv", mode = "wb")
```

# Intro to R

<div class="alert alert-info">
<strong>Go to slides</strong> 
</div>

## Create objects in R

1. R As Calculator
```{r}
3+5  
12/7  
12*7
3^2 # 3 to the power of 2

```

2. Assigning value to an object 
  - Word *object* and *variable* can differ in meaning, but here used synonymously. 
```{r}
# Creating objects
area_hectares <- 1.0 # asign

area_hectares # call
```

3. Discuss the naming rules and conventions

<div class="alert alert-info">
<strong>Go to slides</strong> 
</div>

    
4. Arithmetic with created objects
```{r}
# Arithmetic with objects

area_hectares * 2.47 

area_hectares <- 2.5  
area_hectares * 2.47  

area_acres <- area_hectares * 2.47 

area_hectares <- 50 
area_acres  

area_hectares * 2.47 

```

5. **Exercise 1 ** 
 
<div class="alert alert-info">
<strong>Go to slides</strong> 
</div>

## Functions 

1. What is a function and the arguments
2. Function returns a value
```{r}
# Functions 

sqrt(9)

```

3. Arguments of the functions 
```{r}
# Function with multiple arguments


round(3.1415)
args(round)

round(3.1415, digits = 2)

# ?round
```

4. **Exercise 2**

<div class="alert alert-info">
<strong>Go to slides</strong> 
</div>


## Vectors and data types

1. Data types 
  - `numeric` (`double`)
  - `character`
  - `logical`
  - `integer`
  - `complex`
  - `raw`

2. Create vectors with `c()` (combine) function
```{r}
#Vectors 
hh_members <- c( 3, 7, 10, 6)
hh_members

respondent_wall_type <- c('muddaub', 'burntbricks', 'sunbricks')
respondent_wall_type
```

3. Inspect content of the vector
```{r}
#Length of vector
length(hh_members)
length(respondent_wall_type)

```

4. Add elements to a vector with `c()`
```{r}
# Adding elements to a vector
possessions <- c('bicycle', 'radio', 'television')
possessions <- c(possessions, 'mobile_phone') # add to the end of the vector
possessions <- c('car', possessions) # add to the beginning of the vector

possessions

```


5. **Exercise 3**

<div class="alert alert-info">
<strong>Go to slides</strong> 
</div>


6. OPTIONAL **Exercise 4**

<div class="alert alert-info">
<strong>Go to slides</strong> 
</div>


7. Subsetting vectors 

```{r}
# Subsetting vectors
respondent_wall_type <- c('muddaub', 'burntbricks', 'sunbricks')

# Subsetting by index
respondent_wall_type[2] # 2 element of the vector
respondent_wall_type[c(3,2)] # 3 and 2 element of the vector

# Subsetting by logical vector
hh_members <- c(3, 7, 10, 6)
hh_members[c(TRUE, FALSE, TRUE, TRUE)] # 1,3 and 4 element of the vector 

# Subsetting by condition

hh_members > 5   # condition creates a vector of logical values
hh_members[hh_members > 5]
hh_members[hh_members < 4 | hh_members > 7]
hh_members[hh_members >= 4 & hh_members <= 7]

possessions <- c("car", "bicycle", "radio", "television", "mobile_phone")

possessions[possessions == "car" | possessions == "bicycle"] 

possessions %in% c("car", "bicycle")
possessions[possessions %in% c("car", "bicycle")]


```

## Missing Data 
1. Missing data notation in R - NA
2. Use of `na.rm = TRUE` argument
```{r}
#Missing values

rooms <- c(2, 1, 1, NA, 7)
mean(rooms)
max(rooms)

mean(rooms, na.rm = TRUE)
max(rooms, na.rm = TRUE)
```

3. Useful functions: `is.na()`, `na.omit()`
```{r}
rooms
is.na(rooms)
!is.na(rooms)
rooms[!is.na(rooms)]

```

4. **Exercise 5**

<div class="alert alert-info">
<strong>Go to slides</strong> 
</div>
